Executive Constraints and Civil Conflict Onset

### 2020-08-29 paper.Rmd
Manuscript edits

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 13052
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: Lorem ipsum
---

@@

The primary onset dates were determined according to UCDP/PRIO's
episodic definition where a new onset is coded for the same
incompatibility when there is at least a one year pause in the
fighting or drop below the battle death threshold. The primary
rationale is that civil conflicts often go through several distinct
phases of fighting which can make it difficult to accurately code the
end of a conflict. The episodic definition avoids potentially
conflating briefs lulls involving the same rebel groups and leaders
versus the re-emergence of previously long dormant
conflicts. Furthermore, in both scenarios, the theory linking
executive constraints to the incentives for political violence should
still hold.[^robust]

[^robust]: Due to the potential sensitivity of results to changing
    conflict definitions [@Sambanis2004b], all analyses were also run
    using only unique onsets as well as major episodic onsets with a
    threshold of 1000 battle deaths. The results are available in
    [Appendix C](#appendix-c).

@@

(ref:confusion) Confusion matrix with a 0.168 cutoff

@@

# Discussion {-}

>>>>

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 14000
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: Lorem ipsum
---

@@

The primary onset dates were determined according to UCDP/PRIO's
episodic definition where a new onset is coded for the same
incompatibility when there is at least a one year pause in the
fighting or drop below the battle death threshold. The primary
rationale is that civil conflicts often go through several distinct
phases of fighting which can make it difficult to accurately code the
end of a conflict. The episodic definition avoids potentially
conflating briefs lulls involving the same rebel groups and leaders
versus the re-emergence of previously long dormant
conflicts. Furthermore, in both scenarios, the theory linking
executive constraints to the incentives for political violence should
still hold.

@@

(ref:confusion) Confusion matrix with a 0.168 cutoff for the median
    posterior predicted probabilities.

@@

## Alternative Specifications

An additional issue which may affect the validity of the model is the
potential sensitivity of the results to changing operational
definitions of civil conflict. Operationalizing civil conflict onset
is a difficult endeavour [@Sambanis2004b]. For example, when exactly
does a conflict start? Is it at the initial formation of a rebel
group? The first battle death? Or, when a certain threshold in
conflict intensity has been achieved? Similar questions can be raised
about when a conflict ends and how to clearly distinguish conflict
termination from brief lulls in fighting. Different operational
definitions can affect estimates and hinder the ability to compare
results across studies. While datasets such as UCDP/PRIO have done
much to standardise the operational definition of civil conflict,
there is still a significant degree of researcher freedom in how onset
is included as the outcome variable in a model.

For the results presented so far the outcome variable has followed the
episodic definition of civil conflict onset from UCDP/PRIO --- 25
battle death threshold with new episodes coded after one year of
inactivity --- which captures both high and low intensity civil
conflicts. How robust are these findings when only examining
full-scale civil wars? There is no *a prior* theoretical reason to
expect that the relationship between executive constraints and civil
conflict onset should not be negative when only examining civil
wars. In fact, at least one of the mentioned theoretical mechanisms,
namely that executive constraints may affect the ability of rebel
groups to mobilize potential fighters by offering local security
guarantees against government repression, posited that the theory may
*not* hold for low intensity conflicts featuring predatory rebel
groups. Thus, if the relationship were not to be found when re-running
the model with only high intensity conflict onsets that would be a
significant challenge to the external validity of the initial results.

In the UCDP/PRIO armed conflict dataset, high intensity is defined at
the threshold of 1,000 battle deaths. High intensity conflicts, or
civil wars, can be coded according to when the threshold for high
intensity is crosses either by a single year or by the cumulative
battle deaths.[^battle_deaths] I therefore re-run the same model,
including the same priors and runtime parameters, twice for the two
different operational definitions. In both cases, I retain an episodic
delineation between separate conflicts --- in other words, conflicts
over the same incompatibility can have multiple onsets when there are
lulls in fighting.[^episode_gaps] For the former, the onset year is
set as the first year which crosses the high intensity threshold for
each period of sustained fighting of at least 25 battle deaths. For
the latter, onset is the coded conflict year for the cumulative period
passing the high intensity threshold.

[^battle_deaths]: A more robust modeling strategy would be to directly
    predict the number of battle deaths and then apply *a posteriori*
    decision criteria to the predicted results to classify
    country-years as low-intensity or high intensity conflict. This
    would avoid the arbitrary nature of battle deaths thresholds for
    determining the intensity types; however, battle death data tends
    to be limited in time span and there would be the additional issue
    of data quality.

[^episode_gaps]: Another robustness check would be to re-run the model
    for different episode definitions based on varying intervals of
    fighting inactivity. For example, coding termination after one,
    two, or three years of inactivity. This remains outside the scope
    of the present paper, however, due to lack of computational
    resources for such a high number of model runs.

(ref:overlay) Comparison of regression slopes between the original
    model run, "Low & High Intensity", and the model runs for the two
    different operational definitions of high intensity
    conflict. Credible intervals are 80 percent posterior quantile
    intervals.

```{r fig.width = 8, fig.height = 4, fig.cap = "\\label{fig:overlay} (ref:overlay)"}
models <- c("full_model", "major_only", "cumulative_only")
ll <- lapply(models, function(model) {
    post <- file.path("posteriors", model, "reg_posteriors.csv") %>%
        read_post
})

args <- list(pars = "beta",
             legend_labels = c("Low & High Intensity",
                               "High Intensity",
                               "Cumulatively High Intensity"))
do.call(overlay_pars, c(ll, args)) +
    scale_x_discrete(labels = colnames(beta))
```

```{r}
ix_probs <- lapply(seq_along(models), function(i) {
    beta <- extract(ll[[i]], "beta")

    # Really could just grab this from the existing `beta` variable,
    # but just to be safe...
    load(file.path("posteriors", models[i], "input.RData"),
         envir = e <- new.env())
    colnames(beta) <- c("exec_constraints", colnames(e$X),
                        sprintf("exec * %s",
                                colnames(e$X)[e$interaction_idx]))

    mean(beta[, "exec * rgdpepc"] < 0)
})

ix_probs <- unlist(ix_probs)
names(ix_probs) <- models
```

Figure \ref{fig:overlay} plots the original results compared to the
results from running the model with the two different operational
definitions of high intensity conflict.[^theta] The majority of
depicted parameter estimates are fairly stable across runs. Notably, the relationship between executive constraints and conflict onsets remains negative with a high posterior probability of being less than zero. The estimates for the interaction effect between executive constraints and GDP per capita, however, become far to uncertain to be able to conclusively assert an association. The probability that the interaction term is less than zero for the high intensity model run, *i.e.* where at least a single year crosses the high intensity threshold, is only `r fmt(ix_probs["major_only"])` while the probability for the cumulative intensity model is only
`r fmt(ix_probs["cumulative_only"])`.

[^theta]: Since the measurement model for executive constraints is
    embedded within the larger regression predicting civil conflict
    onset, the latent factor $\theta$ is re-estimated for each model
    run. The stochastic nature of posterior sampling ensures that the
    estimates are never be exactly the same; however, the differences
    are inconsequential. In fact the correlation between the main
    model and both the cumulatively high intensity model and the major
    conflicts only model is effectively one.

(ref:confusion_major) Confusion matrix for conflicts with at least a
    single year passing the high intensity threshold with a 0.168
    cutoff for the median posterior predicted probabilities.

```{r confusion_major}
extra <- read_post("./posteriors/major_only/extra_posteriors.csv")

p_hat <- extract(extra, "p_hat")
p_median <- apply(p_hat, 2, median)

load("./posteriors/major_only/input.RData")

m <- table(p_median > .168, final.df$lmajor_onset)
kable(m, "latex", align = "c", booktabs = T,
      linesep = "", escape = F, caption = "hello")
```

(ref:confusion_cumulative) Confusion matrix for conflicts with a
    cumulative high intensity with a 0.168 cutoff for the median
    posterior predicted probabilities.

```{r confusion_cumulative}
extra <- read_post("./posteriors/cumulative_only/extra_posteriors.csv")

p_hat <- extract(extra, "p_hat")
p_median <- apply(p_hat, 2, median)

load("./posteriors/cumulative_only/input.RData")

m <- table(p_median > .168, final.df$lcum_onset)
kable(m, "latex", align = "c", booktabs = T,
      linesep = "", escape = F,
      caption = "world")
```

How well does the model fit the data when only examining high intensity conflicts?

# Conclusion {-}
